Constructing Sailing Match Race Schedules: Round-Robin Pairing Lists

### 2015-04-13 paper.tex
Mucking around

\title{Constructing Sailing Match Race Schedules: Round Robin Pairing Lists}
\titlerunning{Match Race Scheduling}
%\toctitle{}
\author{Craig Macdonald \and Ciaran McCreesh \and Alice Miller \and Patrick Prosser}
\institute{University of Glasgow, Glasgow, Scotland \\ \email{patrick.prosser@glasgow.ac.uk}}
\authorrunning{Macdonald, McCreesh, Miller, Prosser}
\maketitle

@@

This work describes a round robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on their own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimised. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from ISAF definition of optimality.

This paper describes the scheduling problem. We present a constraint programming solution and
document how this model was constructed incrementally. We believe this illustrates how adaptable
constraint programming is, i.e.\ given a robust model this can be incrementally developed to
address all of the required criteria. Our schedules are then constructed in a sequence of stages,
not dissimilar to the hybrid and decomposition based approaches in \cite{lombardi2012}. Some ISAF
published schedules are presented along with some of our own, some of these being new.  These
schedules are essentially blank forms, made available to event organisers, which can be filled in
with names of actual competitors, i.e.\ these schedules are reusable.

@@

\section{Problem Definition: Round Robin Pairing Lists}\label{sec:definition}

The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the
competing skippers are scheduled into flights and matches, known as pairing lists. Several of these
criteria are applicable when the number of skippers exceeds the number of available boats, in which
case skippers have to change in and out of boats, and are allowed time in the schedule to check and
fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each
match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches.
This results in 9 flights each of 5 matches.  Each flight takes about 50 minutes, consequently eight
or nine flights a day is a reasonable achievement for most events~\cite{isaf}. Consequently, the
number of boats and skippers involved is typically small.

The criteria for  match racing schedules (pairing lists) are given in ISAF ``International Umpires'
and Match Racing Manual'' \cite{isaf}, section M.2 Recommended Criteria for Round Robin Pairing
Lists, and are detailed below.

\begin{framed}
\noindent Principal Criteria in Order of Priority:
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item *When skippers have an even number of matches, they have the same number of port and starboard
    assignments.
\item *When skippers have an odd number of matches, the first half of the skippers will have one
    more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of the next flight.
\item No skipper should have more than two consecutive port or starboard assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as equally as possible.
\item In flights with five or more matches, no skipper should be in the next-to-last match in a
    flight and then in the first match of the next flight.
\item If possible, a skipper should be starboard when meeting the nearest lowest ranked skipper
    (i.e.\ \#1 will be starboard against \#2, \#3 will be starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria~10, 11 and~12 only apply when there are fewer boats than skippers; 11 and~12
override~6 when changes are required, and~13 applies when there are more boats than skippers.
\end{framed}

We have rephrased criteria~2 and~3 (denoted *) to clarify an error in \cite{isaf}. Note that the
order of matches within a flight is significant and is constrained, as is the order of flights
within a schedule and the position of skippers within a match. This permits fair schedules that
provide the sufficient changeover time for skippers changing boats, etc.

@@

\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has a temporal view of the schedule
in the array of constrained integer variables $\timeSlot[0]$ to $\timeSlot[{t-1}]$, and a state view
in the array of constraint integer variables $\state[0]$ to $\state[{f-1}]$. The state of this
skipper in flight $i$ is found in $\state[i]$, and this corresponds to time slots $\timeSlot[{i \cdot
m}]$ to $\timeSlot[{m(i+1)-1}]$.

\begin{figure}[p]
\setlength{\mathindent}{1em}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
\setlength{\abovecaptionskip}{0pt}
\begin{framed}
\begin{align}
    \intertext{A copy of these variables and constraints is created for each skipper $\sigma$:\smallskip}
    &\forall \tau \in \{0 \nldots t{-}1\}: \timeSlot[{\tau}] \in \{{-}2 \nldots t{-}1\} \setminus \{\sigma\} \label{var:timeSlot} \tag{V1} \\
    &\forall i \in \{0 \nldots f{-}1\}: \state[{i}] \in \{\FIRST,\MID,\LAST,\BYE,\END\} \label{var:state} \tag{V2} \\
    &\forall i \in \{0 \nldots f\}: \label{C1} \tag{C1} \\
    &\begin{alignedat}{3}
    &\hspace*{1em} \state[{i}] = \FIRST &\ &\Leftrightarrow&&\ \timeSlot[{m \cdot i}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \MID   &\ &\Leftrightarrow&&\ \exists j \in \{m \cdot i + 1 \nldots m \cdot (i + 1) - 2\} : \timeSlot[{j}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \LAST  &\ &\Leftrightarrow&&\ \timeSlot[m \cdot (i + 1) - 1] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \BYE   &\ &\Leftrightarrow&&\ \forall j \in \{m \cdot i \nldots m \cdot (i + 1) - 1\} : \timeSlot[j] = -1 \\
    &\hspace*{1em} \state[{i}] = \END   &\ &\Leftrightarrow&&\ \forall j \in \{m \cdot i \nldots m \cdot (i + 1) - 1\} : \timeSlot[j] = -2
    \end{alignedat} \nonumber \\
    &\eachOccursExactlyOnce(\timeSlot, \{0, \nldots, n{-}1\} \setminus \{\sigma\}) \label{C2} \tag{C2} \\
    &\validStateSequence(\state) \label{C3} \tag{C3} \\[0.1cm]
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] \in \{0,1\} \label{V3} \tag{V3} \\
    & \totalChanges \in \mathbb{N} \label{V4} \tag{V4} \\
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] = 1 \Leftrightarrow \state[{i}] = \BYE \wedge \state[{i + 1}] \neq \BYE \label {C4} \tag{C4} \\
    & \totalChanges = \textstyle\sum \change \label{C5} \tag{C5}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{Match and temporal perspectives:\smallskip}
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} : \nonumber \\
    & \hspace*{1em} \match[{i,j}] \in \{0..t{-}1\} \tag{V5} \label{V5} \\
    & \hspace*{1em} \match[{j,i}] \equiv \match[{i,j}] \nonumber \\
    & \hspace*{1em} \forall k \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{2em} \match[{i,j}] = k \Leftrightarrow \sigma[{i}].\timeSlot[{k}] = j \wedge \sigma[{j}].\timeSlot[{k}] = i \tag{C6} \label{C6} \\[0.1cm]
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} \nonumber : \\
    & \hspace*{1em} \modMatch[{i,j}] \in \{0..f{-}1\} \label{V6} \tag{V6} \\
    & \hspace*{1em} \modMatch[{j,i}] \equiv \modMatch[{i,j}] \nonumber  \\
    & \hspace*{1em} \modMatch[{i,j}] = \match[{i,j}] / m \tag{C7} \label{C7} \\
    & \forall i \in \{0 \nldots n{-}1 \} : \allDifferent(\modMatch[{i}]) \tag{C8} \label{C8} \\[0.1cm]
    & \forall \tau \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{1em} \timeVar[{\tau}] \in \{(0,1) \nldots (n{-}2,n{-}1)\} \tag{V7} \label{V7} \\
    & \hspace*{1em} \forall i \in \{0 \nldots n{-}2\} : \forall  j \in \{ i{+}1 \nldots n{-}1 \} : \timeVar[{\tau}] = (i,j) \Leftrightarrow \match[{i,j}] = \tau \tag{C9} \label{C9} \\[0.1cm]
    & \totalBoatChanges = \textstyle\sum \sigma.\totalChanges \label{V8} \tag{V8} \\
    & \minimise(\totalBoatChanges) \label{C10} \tag{C10}
\end{align}
\end{framed}
\caption{Our constraint model, from a skipper perspective (top) and a match and temporal perspective
(below). The number of skippers is $n$, and $m$ is the number of matches in a flight. The number of
flights is $f = \lceil n(n-1)/m \rceil$, and there are $t = n(n-1)/2$ matches in
total.}\label{model:stage1}
\end{figure}

In \eqref{var:timeSlot} the array $\timeSlot$ is defined and in \eqref{var:state}, $\state$ is
defined. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a match with
skipper~$k$ at time~$i$. Variable $\state[{i}]$ gives the state of this skipper in flight~$i$. The
cases in \eqref{C1} state that a skipper can be in a match in the first time slot in the flight, or in
the middle of the flight\footnote{i.e.\ not first and not last.}, or in the last time
slot of the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper
is in a bye (i.e. not in a match in this flight), and if all time slots in a flight are equal to $-2$ then the skipper has finished all matches.

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i+1$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a possible transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished all matches in flight $i$ (i.e.\ is in state $\END$) then that skipper will also have
finished all matches in flight $i+1$ (and so on).

@@

\bigskip \noindent Figure~\ref{schedChanges} gives a pictorial view of the entire model, using the 7
skipper and 6 boat problem. On the right we have the 7 skippers with their states and time slots.
Again, top right, we give the schedule actually produced.  Middle left we have the $\match$ array
and at the bottom the $\timeVar$ array. Above the $match$ array we have the $\modMatch$ array. The
yellow arrows show the channeling between parts of the model.

@@

\afterpage{\clearpage\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{tikzpicture}
        \begin{scope}
            \node (Time) at (-2, 0) { $\timeVar$ };
            \node (Time0) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (0, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (0,1)
                \nodepart[text width=1.5em, align=center] {two}   (2,3)
                \nodepart[text width=1.5em, align=center] {three} (4,5)
            };
            \node (Time1) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (2.2, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (0,2)
                \nodepart[text width=1.5em, align=center] {two}   (4,6)
                \nodepart[text width=1.5em, align=center] {three} (1,5)
            };
            \node (Time2) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (4.4, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (2,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,5)
                \nodepart[text width=1.5em, align=center] {three} (1,3)
            };
            \node (Time3) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (6.6, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (5,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,3)
                \nodepart[text width=1.5em, align=center] {three} (1,4)
            };
            \node (Time4) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (8.8, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,5)
                \nodepart[text width=1.5em, align=center] {two}   (1,6)
                \nodepart[text width=1.5em, align=center] {three} (2,4)
            };
            \node (Time5) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (11.0, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,4)
                \nodepart[text width=1.5em, align=center] {three} (2,5)
            };
            \node (Time6) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (13.2, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,4)
                \nodepart[text width=1.5em, align=center] {two}   (1,2)
                \nodepart[text width=1.5em, align=center] {three} (0,6)
            };
        \end{scope}

@@

    \end{tikzpicture}
    \caption{A pictorial representation of the entire model of the schedule for 7 skippers and 6 boats.
        The schedule is reproduced on the right. In the centre we have the 7 skippers, on left the
        $\modMatch$ and $\match$ arrays, and along the bottom the $\timeVar$
    array. The arrows signify channeling between parts of the model.}
    \label{schedChanges}
\end{figure}
\end{landscape}}

@@

\begin{figure}[tb]
\setlength{\mathindent}{1em}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
\setlength{\abovecaptionskip}{0pt}
\begin{framed}
\begin{align}
    \intertext{The objective value from stage 1 is used as a hard constraint:\smallskip}
    & \totalBoatChanges \leq \beta \tag{C11} \label{C11}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{A copy of these variables and constraints is created for each skipper $\sigma$:\smallskip}
    & \forall i \in \{ 0 \nldots m{-}1 \} : \forall j \in \{ 0 \nldots f{-}1 \} : \nonumber \\
    & \hspace*{1em} \position[{i,j}] \in \{0,1\} \tag{V9} \label{V9} \\
    & \hspace*{1em} \position[{i,j}] = 1 \Leftrightarrow \timeSlot[{m.j+i}] \geq 0 \tag{C12} \label{C12} \\
    & \forall i \in \{ 0 \nldots m{-}1 \} : \nonumber \\
    & \hspace*{1em} \imbalance[{i}] \in \mathbb{N} \label{V10} \tag{V10} \\
    & \hspace*{1em} \imbalance[{i}] = \abs*{\frac{n-1}{m} - \sum_{j=0}^{f-1} \position[{i,j}]} \label{C13} \tag{C13} \\
    & \maxImbalance \in \mathbb{N} \label{V11} \tag{V11} \\
    & \maxImbalance = \maximum(imbalance) \label{C14} \tag{C14}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{We minimise the maximum imbalance over all skippers:\smallskip}
    & \forall i \in \{ 0 \nldots n{-}1 \} : \imbalance[{i}] \equiv \sigma[{i}].\maxImbalance \label{V12} \tag{V12} \\
    & \maxImbalance \in \mathbb{N} \label{V13} \tag{V13} \\
    & \maxImbalance = \maximum(\imbalance) \label{C15} \tag{C15} \\
    & \minimise(\maxImbalance) \label{C16} \tag{C16}
\end{align}
\end{framed}
\caption{Additions to the constraint model in Figure~\ref{model:stage1} for stage 2. The constant
$\beta$ is the minimum number of boat changes found in stage 1.}\label{model:stage2}
\end{figure}

>>>>

\title{Constructing Sailing Match Race Schedules: Round-Robin Pairing Lists}
\titlerunning{Match Race Scheduling}
%\toctitle{}
\author{Craig Macdonald \and Ciaran McCreesh \and Alice Miller \and Patrick Prosser}
\institute{University of Glasgow, Glasgow, Scotland \\ \email{patrick.prosser@glasgow.ac.uk}}
\authorrunning{Macdonald, McCreesh, Miller, Prosser}
\maketitle

@@

This work describes a round-robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on their own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimised. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from the ISAF definition of optimality.

This paper describes the scheduling problem. We present a constraint programming solution and
document how this model was constructed incrementally. We believe this illustrates how adaptable
constraint programming is: a robust model can be developed incrementally to address all of the
required criteria. Our schedules are then constructed in a sequence of stages, not dissimilar to the
hybrid and decomposition based approaches by Lombardi and Milano \cite{lombardi2012}. Some ISAF
published schedules are presented along with some of our own, some of these being new.  These
schedules are essentially blank forms, made available to event organisers, which can be filled in
with names of actual competitors, i.e.\ these schedules are reusable.

@@

\section{Problem Definition: Round-Robin Pairing Lists}\label{sec:definition}

The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the
competing skippers are scheduled into flights and matches, known as pairing lists. Several of these
criteria are applicable when the number of skippers exceeds the number of available boats, in which
case skippers have to change in and out of boats, and are allowed time in the schedule to check and
fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each
match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches.
This results in 9 flights each of 5 matches.  Each flight takes about 50 minutes, so eight or nine
flights a day is a reasonable achievement for most events~\cite{isaf}. Consequently, the number of
boats and skippers involved is typically small.

The criteria for  match racing schedules (pairing lists) are given in ISAF ``International Umpires'
and Match Racing Manual'' \cite{isaf}, section M.2 ``Recommended Criteria for Round Robin Pairing
Lists'', and are detailed below.

\begin{framed}
\noindent Principal Criteria in Order of Priority:
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item[2*.] When skippers have an even number of matches, they have the same number of port and starboard
    assignments.
\item[3*.] When skippers have an odd number of matches, the first half of the skippers will have one
    more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of the next flight.
\item No skipper should have more than two consecutive port or starboard assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as equally as possible.
\item In flights with five or more matches, no skipper should be in the next-to-last match in a
    flight and then in the first match of the next flight.
\item If possible, a skipper should be starboard when meeting the nearest lowest ranked skipper
    (i.e.\ \#1 will be starboard against \#2, \#3 will be starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria~10, 11 and~12 only apply when there are fewer boats than skippers; 11 and~12
override~6 when changes are required, and~13 applies when there are more boats than skippers.
\end{framed}

We have rephrased criteria~2 and~3 (denoted *) to clarify an error in the manual \cite{isaf}. Note
that the order of matches within a flight is significant and is constrained, as is the order of
flights within a schedule and the position of skippers within a match. This permits fair schedules
that provide the sufficient changeover time for skippers changing boats, etc.

@@

\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has both a temporal view of their
schedule (``who, if anyone, am I racing in the match at time $t$?''), and a state view (``where am
I racing within flight $f$?'').  The temporal view is via the array $\timeSlot$ defined in
\eqref{var:timeSlot}. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a
match with skipper~$k$ at time~$i$.

Variable $\state[{i}]$ \eqref{var:state} gives the state of this skipper in flight~$i$,
corresponding to time slots $\timeSlot[{m \cdot i}]$ to $\timeSlot[{m(i+1)-1}]$.
The cases in \eqref{C1} state that a skipper can be in a match in the first time slot in the flight,
or in the middle of the flight\footnote{i.e.\ not first and not last.}, or in the last time slot of
the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper is in a
bye (i.e. not in a match in this flight), and if all time slots in a flight are equal to $-2$ then
the skipper has finished all matches.

\begin{figure}[p]
\setlength{\mathindent}{1em}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
\setlength{\abovecaptionskip}{0pt}
\begin{framed}
\begin{align}
    \intertext{A copy of these variables and constraints is created for each skipper $\sigma$:\smallskip}
    &\forall \tau \in \{0 \nldots t{-}1\}: \timeSlot[{\tau}] \in \{{-}2 \nldots t{-}1\} \setminus \{\sigma\} \label{var:timeSlot} \tag{V1} \\
    &\forall i \in \{0 \nldots f{-}1\}: \state[{i}] \in \{\FIRST,\MID,\LAST,\BYE,\END\} \label{var:state} \tag{V2} \\
    &\forall i \in \{0 \nldots f\}: \label{C1} \tag{C1} \\
    &\begin{alignedat}{3}
    &\hspace*{1em} \state[{i}] = \FIRST &\ &\Leftrightarrow&&\ \timeSlot[{m \cdot i}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \MID   &\ &\Leftrightarrow&&\ \exists j \in \{m \cdot i + 1 \nldots m \cdot (i + 1) - 2\} : \timeSlot[{j}] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \LAST  &\ &\Leftrightarrow&&\ \timeSlot[m \cdot (i + 1) - 1] \geq 0 \\
    &\hspace*{1em} \state[{i}] = \BYE   &\ &\Leftrightarrow&&\ \forall j \in \{m \cdot i \nldots m \cdot (i + 1) - 1\} : \timeSlot[j] = -1 \\
    &\hspace*{1em} \state[{i}] = \END   &\ &\Leftrightarrow&&\ \forall j \in \{m \cdot i \nldots m \cdot (i + 1) - 1\} : \timeSlot[j] = -2
    \end{alignedat} \nonumber \\
    &\eachOccursExactlyOnce(\timeSlot, \{0, \nldots, n{-}1\} \setminus \{\sigma\}) \label{C2} \tag{C2} \\
    &\validStateSequence(\state) \label{C3} \tag{C3} \\[0.1cm]
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] \in \{0,1\} \label{V3} \tag{V3} \\
    & \totalChanges \in \mathbb{N} \label{V4} \tag{V4} \\
    & \forall i \in \{0 \nldots f{-}2\}: \change[{i}] = 1 \Leftrightarrow \state[{i}] = \BYE \wedge \state[{i + 1}] \neq \BYE \label {C4} \tag{C4} \\
    & \totalChanges = \textstyle\sum \change \label{C5} \tag{C5}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{Match and temporal perspectives:\smallskip}
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} : \nonumber \\
    & \hspace*{1em} \match[{i,j}] \in \{0 \nldots t{-}1\} \tag{V5} \label{V5} \\
    & \hspace*{1em} \match[{j,i}] \equiv \match[{i,j}] \nonumber \\
    & \hspace*{1em} \forall k \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{2em} \match[{i,j}] = k \Leftrightarrow \sigma[{i}].\timeSlot[{k}] = j \wedge \sigma[{j}].\timeSlot[{k}] = i \tag{C6} \label{C6} \\[0.1cm]
    & \forall i \in \{0 \nldots n{-}2\} : \forall j \in \{ i{+}1 \nldots n{-}1 \} \nonumber : \\
    & \hspace*{1em} \modMatch[{i,j}] \in \{0 \nldots f{-}1\} \label{V6} \tag{V6} \\
    & \hspace*{1em} \modMatch[{j,i}] \equiv \modMatch[{i,j}] \nonumber  \\
    & \hspace*{1em} \modMatch[{i,j}] = \match[{i,j}] / m \tag{C7} \label{C7} \\
    & \forall i \in \{0 \nldots n{-}1 \} : \allDifferent(\modMatch[{i}]) \tag{C8} \label{C8} \\[0.1cm]
    & \forall \tau \in \{0 \nldots t{-}1 \} : \nonumber \\
    & \hspace*{1em} \timeVar[{\tau}] \in \{(0,1) \nldots (n{-}2,n{-}1)\} \tag{V7} \label{V7} \\
    & \hspace*{1em} \forall i \in \{0 \nldots n{-}2\} : \forall  j \in \{ i{+}1 \nldots n{-}1 \} : \timeVar[{\tau}] = (i,j) \Leftrightarrow \match[{i,j}] = \tau \tag{C9} \label{C9} \\[0.1cm]
    & \totalBoatChanges = \textstyle\sum \sigma.\totalChanges \label{V8} \tag{V8} \\
    & \minimise(\totalBoatChanges) \label{C10} \tag{C10}
\end{align}
\end{framed}
\caption{Our constraint model, from a skipper perspective (top) and a match and temporal perspective
(below). The number of skippers is $n$, and $m$ is the number of matches in a flight. The number of
flights is $f = \lceil n(n-1)/m \rceil$, and there are $t = n(n-1)/2$ matches in
total.}\label{model:stage1}
\end{figure}

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i+1$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a possible transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input~$\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished all matches in flight $i$ (i.e.\ is in state $\END$) then that skipper will also have
finished all matches in flight $i+1$ (and so on).

@@

\bigskip \noindent Figure~\ref{schedChanges} gives a pictorial view of the entire model, using the 7
skipper and 6 boat problem. On the right we have the 7 skippers with their states and time slots.
Again, on the right we give the schedule actually produced. On the left we have the $\modMatch$ and
$\match$ arrays, and at the bottom the $\timeVar$ array. The arrows show the channeling between
parts of the model.

@@

\afterpage{\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{tikzpicture}
        \begin{scope}
            \node (Time) at (-2, 0) { $\timeVar$ };
            \node (Time0) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (0, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (0,1)
                \nodepart[text width=1.5em, align=center] {two}   (2,3)
                \nodepart[text width=1.5em, align=center] {three} (4,5)
            };
            \node (Time1) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (2.2, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (0,2)
                \nodepart[text width=1.5em, align=center] {two}   (4,6)
                \nodepart[text width=1.5em, align=center] {three} (1,5)
            };
            \node (Time2) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (4.4, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (2,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,5)
                \nodepart[text width=1.5em, align=center] {three} (1,3)
            };
            \node (Time3) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (6.6, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (5,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,3)
                \nodepart[text width=1.5em, align=center] {three} (1,4)
            };
            \node (Time4) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (8.8, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,5)
                \nodepart[text width=1.5em, align=center] {two}   (1,6)
                \nodepart[text width=1.5em, align=center] {three} (2,4)
            };
            \node (Time5) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (11.0, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,6)
                \nodepart[text width=1.5em, align=center] {two}   (0,4)
                \nodepart[text width=1.5em, align=center] {three} (2,5)
            };
            \node (Time6) [font=\tiny, anchor=center, draw, rectangle split, rectangle split horizontal, rectangle split parts=3] at (13.2, 0) {
                \nodepart[text width=1.5em, align=center] {one}   (3,4)
                \nodepart[text width=1.5em, align=center] {two}   (1,2)
                \nodepart[text width=1.5em, align=center] {three} (0,6)
            };
        \end{scope}

@@

    \end{tikzpicture}
    \caption{A pictorial representation of the entire model of the schedule for 7 skippers and 6 boats.
        The schedule is reproduced on the right. In the centre we have the 7 skippers, on left the
        $\modMatch$ and $\match$ arrays, and along the bottom the $\timeVar$
    array. The arrows signify channeling between parts of the model.}
    \label{schedChanges}
\end{figure}
\end{landscape}

\begin{figure}[p]
\setlength{\mathindent}{1em}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
\setlength{\abovecaptionskip}{0pt}
\begin{framed}
\begin{align}
    \intertext{The objective value from stage 1 is used as a hard constraint:\smallskip}
    & \totalBoatChanges \leq \beta \tag{C11} \label{C11}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{A copy of these variables and constraints is created for each skipper $\sigma$:\smallskip}
    & \forall i \in \{ 0 \nldots m{-}1 \} : \forall j \in \{ 0 \nldots f{-}1 \} : \nonumber \\
    & \hspace*{1em} \position[{i,j}] \in \{0,1\} \tag{V9} \label{V9} \\
    & \hspace*{1em} \position[{i,j}] = 1 \Leftrightarrow \timeSlot[{m.j+i}] \geq 0 \tag{C12} \label{C12} \\
    & \forall i \in \{ 0 \nldots m{-}1 \} : \nonumber \\
    & \hspace*{1em} \imbalance[{i}] \in \mathbb{N} \label{V10} \tag{V10} \\
    & \hspace*{1em} \imbalance[{i}] = \abs*{\frac{n-1}{m} - \sum_{j=0}^{f-1} \position[{i,j}]} \label{C13} \tag{C13} \\
    & \maxImbalance \in \mathbb{N} \label{V11} \tag{V11} \\
    & \maxImbalance = \maximum(imbalance) \label{C14} \tag{C14}
\end{align}
\end{framed}\begin{framed}
\begin{align}
    \intertext{We minimise the maximum imbalance over all skippers:\smallskip}
    & \forall i \in \{ 0 \nldots n{-}1 \} : \imbalance[{i}] \equiv \sigma[{i}].\maxImbalance \label{V12} \tag{V12} \\
    & \maxImbalance \in \mathbb{N} \label{V13} \tag{V13} \\
    & \maxImbalance = \maximum(\imbalance) \label{C15} \tag{C15} \\
    & \minimise(\maxImbalance) \label{C16} \tag{C16}
\end{align}
\end{framed}
\caption{Additions to the constraint model in Figure~\ref{model:stage1} for stage 2. The constant
$\beta$ is the minimum number of boat changes found in stage 1.}\label{model:stage2}
\end{figure}

\begin{SCtable}[1.8][p]
    \setlength{\tabcolsep}{3pt}
    \begin{tabular}{cccc}
        \toprule
        Flight & \multicolumn{3}{c}{Matches} \\ \midrule
        0 & (0,6) & (2,3) & (4,5) \\
        1 & (0,2) & (4,1) & (6,5) \\
        2 & (2,1) & (0,5) & (6,3) \\
        3 & (5,1) & (0,3) & (6,4) \\
        4 & (3,5) & (6,1) & (2,4) \\
        5 & (3,1) & (0,4) & (2,5) \\
        6 & (3,4) & (6,2) & (0,1) \\ \bottomrule
    \end{tabular}
    \caption{A round-robin schedule with 7 skippers, 6 boats and 7 flights, which is the result of
        stages 1, 2, and 3. It has 6 boat changes and imbalance 1. An example of imbalance is
        skipper $\sigma[{0}]$ with 2 matches in position 0 (first), 3 in position 1 and 1 (mid) in
        position 2 (last). Perfect balance would be 2 matches in each position, only achieved by
    skipper 2.}\label{tab2}
\end{SCtable}
\clearpage % I feel oh so very dirty
}

@@
